---
title: "Map of Hawaiian Land Use, Land Cover, and Watersheds"
author: "Nathaniel Burola"
date: "2020-03-04T13:47:08+02:00"
output: html_document
banner: "img/banners/hawaii.jpg"
---

```{r setup, include=FALSE, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hawaiian Land Use, Land Cover, and Watersheds 
This project depicts Hawaiian land use and land cover for all of the main islands in Hawaii. Categories are utilized and displayed on an interactive map such as urban areas, rural areas, forest, sand, rock, etc to show different areas that are covered in different ways. Watershed areas in terms of square mile density are also displayed for all of the main islands in Hawaii. This allows you to see which areas of the main islands in Hawaii are filled to the brink with watershed areas.

### Attaching all relevant packages 
```{r, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}
library(tidyverse)
library(sf)
library(tmap)
library(leaflet)
library(htmlwidgets)
library(htmltools)
```

### Reading in the land use/land cover data 
```{r, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}

hawai_land <- read_sf(dsn = ".", layer = "Land_Use_Land_Cover_LULC")

st_crs(hawai_land) = 4326 #Setting the coordinate reference system (CRS) to 4326 

plot(hawai_land)

```

### Reading in the watershed data 
```{r, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}

hawai_water <- read_sf(dsn = ".", layer = "Watersheds")

st_crs(hawai_water) = 4326 #Setting the CRS to 4326 

plot(hawai_water)
 
```

## Map 1: Landuse and Land Cover Types for All Main Hawaiian Islands 

### Finding out the different types of land use layers in the landcover column 
```{r, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}
unique(hawai_land$landcover) #Informs what types of land use covers exist in the dataset 

#Creating a new column in the data set and categorizing the following land use covers in order to streamline organization 
landmap <- hawai_land %>%  
  mutate(
    landmapnew = case_when(
      landcover == "Cropland and Pasture" ~ "Farmland", 
      landcover == "Commerical and Services" ~ "Urban", 
      landcover == "Residential" ~ "Human Dwellings", 
      landcover == "Evergreen Forest Land" ~ "Forest", 
      landcover == "Other Urban or Built-up Land" ~ "Urban", 
      landcover == "Mixed Rangeland" ~ "Rural",
      landcover == "Industrial" ~ "Urban", 
      landcover == "Streams and Canals" ~ "Watershed", 
      landcover == "Orchards, Groves, Vineyards, Nurseries and Ornamental Horticultural Areas" ~ "Rural", 
      landcover == "Shrub and Brush Rangeland" ~ "Rangeland", 
      landcover == "Forested Wetland" ~ "Forest", 
      landcover == "Reservoirs" ~ "Watershed", 
      landcover == "Nonforested Wetland" ~ "Forest", 
      landcover == "Bare Exposed Rock" ~ "Rock", 
      landcover == "Sandy Areas Other than Beaches" ~ "Sand", 
      landcover == "Transportation, Communications and Utilities" ~ "Urban", 
      landcover == "Herbaceous Rangeland" ~ "Rangeland", 
      landcover == "Beaches" ~ "Beaches", 
      landcover == "Other Agricultural Land" ~ "Farmland", 
      landcover == "Lakes" ~ "Watershed", 
      landcover == "Strip Mines, Quarries, and Gravel Pits" ~ "Mining", 
      landcover == "Mixed Barren Land" ~ "Barren Land", 
      landcover == "Bays and Estuaries" ~ "Watershed",
      landcover == "Mixed Urban or Built-up Land" ~ "Urban",
      landcover == "Transitional Areas" ~ "Unknown", 
      landcover == "Industrial and Commercial Complexes" ~ "Urban", 
      landcover == "Confined Feeding Operations" ~ "Urban", 
      landcover == "0" ~ "NA"
  
    )
  )

tmap_mode("view")
```

### Creating the map which contains all the new categories for the land use layers in Hawaii 
```{r, echo=TRUE, message=FALSE, warning=FALSE}
tm_shape(landmap) + 
  tm_polygons("landmapnew", title = "Land Cover Types for All Main Hawaiian Islands", style = "fixed") + 
  tm_borders() + 
  tm_layout(legend.title.size = 1,
           legend.text.size = 0.6,
          legend.position = c("left","bottom"),
          legend.bg.color = "white",
          legend.bg.alpha = 1) + 
  tm_basemap("Esri.NatGeoWorldMap") 
```

As seen in the land use layer of the Hawaii map, most of the Hawaiian islands are covered in forest followed by a lot of rangeland. There are small pockets of urban areas which makes sense since Hawaii is a tourist destination and a popular place to live. However, it is remarkable to see just how much nature is preserved! 


## Map 2: Different watershed areas for all of the main Hawaiian Islands
```{r, echo=TRUE, message=FALSE, warning=FALSE}
#Constructing the map of all the various watersheds with the Hawaiian Islands 

tm_shape(hawai_water) + 
  tm_polygons("area_sqmi", title = "Square Miles of Watershed Areas") + 
  tm_borders() + 
  tm_layout(legend.title.size = 1,
           legend.text.size = 0.6,
          legend.position = c("left","bottom"),
          legend.bg.color = "white",
          legend.bg.alpha = 1) + 
  tm_basemap("Esri.NatGeoWorldMap")
  


```

As seen in the watershed area map of Hawaii, most if not all the islands are covered by 0 - 4 square miles of watershed areas. There are a couple of watershed areas that are greater than 6 square miles according to the map, however. 
