---
title: "Data Wrangling and Visualization with Snowshoe Hares/Principal Components Analysis (PCA)"
author: "Nathaniel Burola"
date: "2020-03-04T13:47:08+02:00"
output: html_document
banner: "img/banners/snowshoe.png"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

local({
	
	# The directory where Pandoc will be extracted. Feel free
	# to adjust this path as appropriate.
	dir <- "~/rstudio-pandoc"
	
	# The version of Pandoc to be installed.
	version <- "2.7.1"
	
	# Create and move to the requested directory.
	dir.create(dir, showWarnings = FALSE, recursive = TRUE)
	owd <- setwd(dir)
	on.exit(setwd(owd), add = TRUE)
	
	# Construct path to pandoc.
	root <- "https://s3.amazonaws.com/rstudio-buildtools"
	suffix <- sprintf("pandoc-%s-windows-x86_64.zip", version)
	url <- file.path(root, "pandoc-rstudio", version, suffix)
	
	# Download and extract pandoc.
	file <- basename(url)
	utils::download.file(url, destfile = file)
	utils::unzip(file)
	unlink(file)
	
	# Write .Renviron to update the version of Pandoc used.
	entry <- paste("RSTUDIO_PANDOC", shQuote(path.expand(dir)), sep = " = ")
	contents <- if (file.exists("~/.Renviron")) readLines("~/.Renviron")
	filtered <- grep("^RSTUDIO_PANDOC", contents, value = TRUE, invert = TRUE)
	amended <- union(filtered, entry)
	writeLines(amended, "~/.Renviron")
	
	# Report change to the user.
	writeLines("Updated .Renviron:\n")
	writeLines(amended)
	writeLines("\nPlease restart RStudio for these changes to take effect.")
	
})
```

## Data wrangling and visualization with snowshoe hares (Lepus americanus)
Snowshore hares, otherwise known as Lepus americanus, are a keystone prey species that play an ecologically important role in northern boreal forests and experience population fluctuations of 8-11 years. This capture-recapture study of snowshoe hares was conducted at 5 locales in the Tanana valley from study sites Tok in the east to Clear in the west from 1999-2012 with a time duration of 13 years. Snowshoe hare population densities were the highest in 1999 and declined thereafter. This study did not find any declines in apparent survival during declining densities in these study populations.
```{r, echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)
install.packages("here")
install.packages("janitor")
install.packages("kableExtra")
install.packages("naniar")
install.packages("skimr")
install.packages("ggfortify")
install.packages("tidyverse") 
install.packages("hrbrthemes")
install.packages("viridis")
install.packages("magrittr")
install.packages("qwraps2")
install.packages("factoextra")
```

### Attaching all relevant packages
```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(tidyverse)
library(janitor)
library(here)
library(kableExtra)
library(naniar)
library(skimr)
library(ggfortify)
library(tmap)
library(ggplot2)
library(hrbrthemes)
library(qwraps2)
library(magrittr)
library(factoextra)
```

### Reading in the data to form a new dataframe 
```{r, echo = TRUE, message = FALSE, warning = FALSE}
snowshoe <- read_csv("showshoe_lter.csv")
```

### Data exploring and wrangling 
```{r, echo = TRUE, message = FALSE, warning = FALSE, results = 'hide'}
summary(snowshoe) #Gives a summary of how many different variables there are in the dataframe
View(snowshoe) #Opens a new dataframe with the dataset already in it 
names(snowshoe) #Gives the number of column headers in the data set 
head(snowshoe) #Gives a preview of the data set 
tail(snowshoe) #Gives a preview of the data set 

#Initial reaction: Columns that include NA data are the following: time, right ear, sex, age, hindfoot, and notes. 

#What variables may be related to each other with this data? Could include sex, age, and weight. Maybe grid and date? 

#What do you want to graph with the snowshoe data? 

#Statement: I want to see how the weight of all the snowshoe hares both female and male will vary across the years of the study 
```
 
### Filtering and constructing a new dataframe with desired variables 
```{r, echo = TRUE, message = FALSE, warning = FALSE}
snowfiltered = subset(snowshoe, select = -c(time, grid, trap, l_ear, r_ear, hindft, notes, b_key, session_id, study)) #Filtered out all the columns that are not desired in the new dataframe 

#Transforming all of the classes of the variables from character to numeric 
```

### Filtering the dataset 
```{r, echo = TRUE, message = FALSE, warning = FALSE}
#Filtering the dataset in order to get lower case letters for male and females and filtering out the rest of the NA values 
snow_tidy <- snowfiltered %>%  
  filter(sex %in% c ("f", "m", "M", "F")) %>%  
  mutate(
    sex = case_when(
      sex == "f" ~ "F", 
      sex == "m" ~ "M")) %>%  
  filter(sex!= "NA")
```

### Descriptive introductory summary of what the project is all about (3-4 sentences)
```{r, echo = TRUE, message = FALSE, warning = FALSE}
#Snowshore hares, otherwise known as Lepus americanus, are a keystone prey species that play an ecologically important role in northern boreal forests and experience population fluctuations of 8-11 years. This capture-recapture study of snowshoe hares was conducted at 5 locales in the Tanana valley from study sites Tok in the east to Clear in the west from 1999-2012 with a time duration of 13 years. Snowshoe hare population densities were the highest in 1999 and declined thereafter. This study did not find any declines in apparent survival during declining densities in these study populations. 
```

### Finalized graph about the Bonanza Creek snowshoe hare population  
```{r, message = FALSE, warning = FALSE, fig.align="left"}
#Creating a histogram based on the weight of the snowshore hares with a frequency count of the data split between by sex (male/female)
snow_plot <- ggplot(snow_tidy, aes(x=weight, color=sex, fill=sex)) +
  geom_histogram(position="identity", alpha=0.5, brewerPalette="Paired",alpha=0.5) + 
  labs(title="Weight Distribution of Female and Male Snowshoe Hares from 1999 - 2012",x="Weight (g)", y = "Frequency Count of Snowshoe Hares", caption = "Weight distribution of male and female snowshoe hares (Lepus americanus) \n in Bonanza Creek Experimental Forest (1999-2012) \n (Source: Knut, Kielland, et al. “	Snowshoe Hare Physical Data in Bonanza Creek Experimental Forest: 1999-2012.” Bonanza Creek Experimental Forest, 10 Feb. 2017.") + 
  theme_bw()

snow_plot+theme(plot.caption=element_text(hjust=0.5))
```

There is a wide range of weights associated with either gender of the snowshoe hare population in Bonanza Creek from 1999-2012. However, there are noticeable differences between male and female snowshoe hares. It would seem that overall there were more male snowshoe hares that weighed in the 1,100 - 1,400 gram range than compared to female snowshoe hares.

### Creating a new data frame with desirable traits  
```{r, echo = TRUE, message = FALSE, warning = FALSE, results = 'hide'}

#Creating a new data frame to include variables that cannot be analyzed in the numeric class 
snowtable = subset(snowshoe, select = -c(time, grid, trap, l_ear, session_id, study, notes, b_key, age, r_ear, hindft)) 

#Renaming columns in the dataframe to be capitalized 
names(snowtable)[names(snowtable) == "sex"] <- "Sex"
names(snowtable)[names(snowtable) == "date"] <- "Year of Study"
names(snowtable)[names(snowtable) == "weight"] <- "Weight of Hare (g)"


#Filtering out the NA values for the new data frame to only get male and female 
snow_analysis <- snowtable %>%  
  filter(Sex %in% c ("f", "m", "M", "F")) %>%  
  mutate(
    Sex = case_when(
      Sex == "f" ~ "F", 
      Sex == "m" ~ "M")) %>%  
  filter(Sex!= "NA")

#Filtering out female snowshow hares to only get the males for data analysis results  
snow_male <- snow_analysis %>%  
 filter(Sex!= "F")

snow_male

#Filtering out male snowshow hares to only get the females for data analysis results 
snow_female <- snow_analysis %>%  
 filter(Sex!= "M")

snow_female
```

### Testing out various statistical analyses  
```{r, echo = TRUE, message = FALSE, warning = FALSE, results = 'hide'}
#Calculating Descriptive Statistics of the Weight of Female Snowshoe Hares 
mean(snow_female$`Weight of Hare (g)`, na.rm = TRUE)
#1289.994
median(snow_female$`Weight of Hare (g)`, na.rm = TRUE)
#1350 
sd(snow_female$`Weight of Hare (g)`, na.rm = TRUE)
#325.08004 
min(snow_female$`Weight of Hare (g)`, na.rm = TRUE)
#240 
max(snow_female$`Weight of Hare (g)`, na.rm = TRUE)
#2170 
range(snow_female$`Weight of Hare (g)`, na.rm = TRUE)
#1930 

```

### Creating the finalized HTML table 
```{r, echo = TRUE, message = FALSE, warning = FALSE}
snow_tbl <- data.frame(
  Statistics = c("Mean", "Median", "Standard Deviation", "Minimum", "Maximum", "Range"),
  Calculations = c(
    "1,290 ",
    "1,350", 
    "325 ", 
    "240", 
    "2,170", 
    "1,930"
  )
)

kable(snow_tbl, caption = '<b>Female Snowshoe Hare Weight Descriptive Statistics</b>') %>%
  kable_styling(full_width = F) %>% 
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "30em", background = "white") %>%  
  footnote(general = "Descriptive statistics which calculate the mean to range of weights for all female snowshoe hare populations in Bonanza Creek Experimental Forest (1999-2012)")   
```

According to the data, female snowshoe hare populations from 1999-2012 in Bonanza Creek have a mean weight of 1,290 grams and a median weight of 1,350 grams. The lightest female snowshoe hare weighed 240 grams while the heaviest female snowshoe hare weighed 2,170 grams. Taking into account the entire female population, there was a range of 1,930 grams from the lightest to the heaviest female snowshoe hare. The standard deviation of the female snowshoe population was 325 grams in relation to the mean weight of 1,290 grams. 

## Principal components analysis (PCA)
What humans eat in terms of their food supply and the scientiifc understanding of relationships between dietary intakes and health have evolved over time. As the United States Department of Agriculture (USDA), this organizations is responsible for maintaining food composition data resources that can meet the needs of diverse users from researchers to health professionals. There is a need for transparent and easily accessible  information about nutrients and a new approach to presenting food profile information as the food industry changes over time. This dataset, titled FoodData Central, presents food profile information in a scientifically rigorous way (Source: FoodData Central . “FoodData Central .” FoodData Central, 2020, fdc.nal.usda.gov/about-us.html.)

### Reading in the dataset 
```{r, echo = TRUE, message = FALSE, warning = FALSE}
nutrients <- read_csv("usda_nutrients.csv")
```

### Doing some data wrangling to clean up and filter the data 
```{r, echo = TRUE, message = FALSE, warning = FALSE}

#Filtered the dataset through food group in onder to have vegetables and vegetable products as well as fruits and fruit juices 

#Retaining observations where the ShortDescrip column would contain the word "RAW" 
usda <- nutrients %>% 
  filter(FoodGroup  == "Vegetables and Vegetable Products" |
           FoodGroup == "Fruits and Fruit Juices") %>%
  filter(str_detect(ShortDescrip, pattern="RAW"))
```

### Useful descrptive summary (3-4 lines)
```{r, echo = TRUE, message = FALSE, warning = FALSE}
#What humans eat in terms of their food supply and the scientiifc understanding of relationships between dietary intakes and health have evolved over time. As the United States Department of Agriculture (USDA), this organizations is responsible for maintaining food composition data resources that can meet the needs of diverse users from researchers to health professionals. There is a need for transparent and easily accessible  information about nutrients and a new approach to presenting food profile information as the food industry changes over time. This dataset, titled FoodData Central, presents food profile information in a scientifically rigorous way (Source: FoodData Central . “FoodData Central .” FoodData Central, 2020, fdc.nal.usda.gov/about-us.html.) 

```

### Creating the summary table for the PCA analysis
```{r, message = FALSE, warning = FALSE, echo=FALSE}

usda_pca <- prcomp(nutrients[9:30], scale = TRUE)

summary(usda_pca)
``` 

### Creating the finalized PCA plot
```{r, PCA, biplot, message = FALSE, warning = FALSE, echo=TRUE}
usda_bilplot <- autoplot(usda_pca, 
                       colour = NA, 
                       loadings.label = TRUE, 
                       loadings.label.size = 3, 
                       loadings.label.colour = "black", 
                       loadings.label.repel = TRUE) + 
   labs(title = "PCA Biplot: Fruit and Vegetable Nutrient Levels",
       x = "PC1 representing 24.80% of variance",
       y = "PC2 representing 11.47% of variance",
       caption = "PCA biplot representing fruit and vegetable nutrient levels \n including variables from protein (g) to zinc levels \n (Source: Food Data Central .“FoodData Central .” FoodData Central, 2020, fdc.nal.usda.gov/about-us.html.") + 
  theme_bw()

usda_bilplot + theme(plot.caption=element_text(hjust=0.5)) + theme(plot.title=element_text(hjust = 0.5))
```

The PCA biplot can be interpreted in multiple trends:

Postive Trend: In terms of postive correlation, Riboflavin and Niacin are postively correlated with one another on account of the length and overlap of their biplot lines. Magnesium and Thiamin are postively correlated with each other as well as Iron and Folate. Sugar and Carbs are postively correlated with one another since they are still in close proximity to one another with their biplot lines.  

Negative Trend: In terms of negative correlation, Sugar and Protein are negatively correlated due to the 180 degree difference between biplot lines earning a correlation of -1. Sugar and Selenium also appear to be negatively correlated as well. 

Neutral Trend: In terms of minimal or no correlation, Sugar and Manganese are minimally correlated with one another due to the 90 degree difference of their biplot lines. Sugar and Fat have minimal or no correlation at all since there is a 270 degree difference between the two biplot lines. 

